Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-08 paper/paper.tex
Algebras and Catamorphisms

\newcommand{\List}[1]{\operator{List}(#1)}
\newcommand{\nil}{\operator{nil}}
\newcommand{\cons}{\operator{cons}}

@@

\subsection{Elements of Category Theory}
A \emph{category}, denoted with $\CatC, \CatD, \CatE$,  is a collection of
objects $A, B, C$, denoted by $\Obj(\CatC)$ and a collection of arrows $f, g, h$
between these objects, denoted by $\Arr{\CatC}$, such that there always exists
an identity arrow $id_{A} : A \to A$ for each object $A$ and for two arrows
$A \xrightarrow{f} B$ and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$\footnote[1]{For
presentation purposes we shall not deal with size issues and assume all the
categories are small.}. The initial object in a category, denoted by $0$, is the
object such that for any other object $A$ there is a unique arrow
$0 \xrightarrow{!} A$.  Dually, the terminal object, denoted by $1$, is the
object such that for any other object $A$ there is a unique arrow
$A \xrightarrow{!} 1$. Initial and terminal object are unique up-to isomorphism.

@@

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
where $A$ is an object of the category called the \emph{carrier} of the algebra
and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ in
$F\Alg(\CatC)$ are arrows $f : A \to B$ in $\CatC$ such that they respect the
structure of the algebra, that is $f \circ \alpha = \beta \circ F(f)$.  The
initial object in this category is called the \emph{initial $F$-algebra}, that
is the $F$-algebra which has a unique $F$-algebra homomorphism into any other
$F$-algebra. If $(A, \alpha)$ is initial this implies $A$ is the \emph{least
fixed-point} for the functor $F$, i.e. $A \iso FA$ is the least solution of the
domain equation $X \iso F X$ by Lambek's lemma.

For example, for a set $A$ and a given the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$ an algebra is a set $X$ together with an arrow
$\alpha : 1 + A \times X \to X$.in $\Set$ given such a type
\coq{A} we can define the monoid operations as \coq{u : unit -> A} for unit and
\coq{m : A * A -> A} for multiplication. Notice that \coq{F X = unit + X * X} is
a functor on \coq{X} which means we can equivalently describe these two maps as
a single one of type \coq{F X -> X} which we call the algebra for the functor
\coq{F} satisfying the unit and associative laws of the monoid. In this work we
will be using algebras to describe the operations of a data type and so we will
not be needing additional equations on these operations.

Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
\coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
this name shortly. \footnote[1]{To be more precise, initial algebras are
isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

>>>>

\newcommand{\List}[1]{\operator{List}(#1)}
\newcommand{\nil}{\operator{nil}}
\newcommand{\cons}{\operator{cons}}
\newcommand{\foldr}{\operator{foldr}}

@@

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by
$\Arr{\CatC}$, such that there always exists an identity arrow
$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
purposes we shall not deal with size issues and assume all the categories are
small.}. The initial object in a category, denoted by $0$, is the object such
that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
Dually, the terminal object, denoted by $1$, is the object such that for any
other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
terminal object are unique up-to isomorphism.

@@

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
where $A$ is an object of the category called the \emph{carrier} of the algebra
and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$. By the property of initial $F$-algebras, for
any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
pronounced ``\emph{catamorphism}''.

For example, for a set $A$ we define the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
structure map $[base, ind] : 1 + A \times B \to B$. The initial $F$-algebra is
clearly the set of lists $\List{A}$. The catamorphism associated with the type
of lists is the unique arrow recursively translating the initial algebra
$[\nil,\cons]$ into the algebra $[base, ind]$:
\begin{align*}
  & \cata{[base,ind]}(\nil) = base\\
  & \cata{[base, ind]}(\cons(a, xs)) = ind(a, \cata{[base, ind]}(xs))
\end{align*}
In functional programming this is commonly referred to as
$\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
$\foldr\;base\;ind = \cata{[base, ind]}$.

\subsection{Dump}
Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
\coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
this name shortly. \footnote[1]{To be more precise, initial algebras are
isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.
